---
title: "FinBIF Last Week"
output: blastula::blastula_email
---

```{r packages, include=FALSE}
library(finbif)
library(lubridate)
library(magick)
library(gt)
library(scales)
library(dplyr)
library(ggplot2)
```

```{r new-obs, include = FALSE}
permissive <- list(
  quality_issues = "both",
  record_reliability = c("reliable", "unassessed", "unreliable"),
  record_quality = c(
    "expert_verified", "community_verified", "unassessed", "uncertain",
    "erroneous"
  )
)

last_week <- list(
  first_import_date_min = today() - 7L,
  first_import_date_max = today() - 1L
)

n <- fb_occurrence(
  filter = c(last_week, permissive),
  count_only = TRUE
)

sources <- finbif:::api_get(
  "sources", list(pageSize = 1000), cache = TRUE
)
sources <- sources$content$results
source_names <- tools::toTitleCase(
  vapply(sources, getElement, character(1L), "name")
)
names(source_names) <-
  paste0("http://tun.fi/", vapply(sources, getElement, character(1L), "id"))
```

# FinBIF Last Week
_Your weekly summary of the latest additions to the Finnish Biodiversity
Information Facility for the week `r format(today()-7, "%A, %B %e")` to
`r format(today()-1, "%A, %B %e, %Y")`_
<br>

****

Last week [__`r n`__](https://laji.fi/en/observation/list?&recordQuality=EXPERT_VERIFIED,COMMUNITY_VERIFIED,NEUTRAL,UNCERTAIN,ERRONEOUS&qualityIssues=BOTH&firstLoadedSameOrAfter=`r as.character(last_week$first_import_date_min)`&firstLoadedSameOrBefore=`r as.character(last_week$first_import_date_max)`) new observations were uploaded to FinBIF

**** 
<br>

## New Observations
```{r by-collection-table, echo = FALSE}
person_responsible <- fb_collections(
  select = person_responsible, supercollections = TRUE
)

n_collections <- fb_occurrence(
  filter = c(last_week, permissive),
  select = c("collection_id"),
  aggregate = c("records", "species"), count_only = TRUE
)

by_collection <- fb_occurrence(
  filter = c(last_week, permissive),
  select = c("collection", "collection_id", "source_id"),
  aggregate = c("records", "species"), n = n_collections
)

by_collection <- mutate(
  by_collection,
  collection = tools::toTitleCase(collection),
  source_id = source_names[source_id],
  collection_qname = gsub("http://tun.fi/", "", collection_id),
  date1 = as.character(last_week$first_import_date_min),
  date2 = as.character(last_week$first_import_date_max)
)

by_collection <- mutate(
  by_collection, person_responsible = person_responsible[collection_qname, ]
)

by_collection <- arrange(by_collection, source_id, collection)

by_collection_tab <- gt(by_collection, groupname_col = "source_id")

by_collection_tab <- summary_rows(
  by_collection_tab,
  groups = TRUE,
  columns = vars(n_records),
  fns = list(`<span style="font-size:10px">Subtotal</span>` = ~sum(.)),
  missing_text = "",
  formatter = function(data, columns = NULL, rows = NULL) {
    fmt(
      data, columns = NULL, rows = NULL,
      fns = function(x) paste0('<span style="font-size:10px">', x, '</span>')
    )
  }
)

by_collection_tab <- grand_summary_rows(
  by_collection_tab,
  columns = vars(n_records),
  fns = list(`<span style="font-size:10px">Total</span>` = ~sum(.)),
  missing_text = "",
  formatter = function(data, columns = NULL, rows = NULL) {
    fmt(
      data, columns = NULL, rows = NULL,
      fns = function(x) paste0('<span style="font-size:10px">', x, '</span>')
    )
  }
)

by_collection_tab <- 
  tab_stubhead(by_collection_tab, label = html("<b>Source</b>"))

by_collection_tab <- cols_merge(
  by_collection_tab,
  vars(n_records, collection_qname, date1, date2),
  pattern = '<span style="font-size:10px"><a href="https://laji.fi/en/observation/list?collectionId={2}&recordQuality=EXPERT_VERIFIED,COMMUNITY_VERIFIED,NEUTRAL,UNCERTAIN,ERRONEOUS&qualityIssues=BOTH&firstLoadedSameOrAfter={3}&firstLoadedSameOrBefore={4}">{1}</a></span>'
)

by_collection_tab <- cols_merge(
  by_collection_tab,
  vars(n_species, collection_qname, date1, date2),
  pattern = '<span style="font-size:10px"><a href="https://laji.fi/en/observation/species?collectionId={2}&recordQuality=EXPERT_VERIFIED,COMMUNITY_VERIFIED,NEUTRAL,UNCERTAIN,ERRONEOUS&qualityIssues=BOTH&firstLoadedSameOrAfter={3}&firstLoadedSameOrBefore={4}">{1}</a></span>'
)

by_collection_tab <- cols_merge(
  by_collection_tab,
  vars(collection, collection_id, person_responsible),
  pattern =
    '<span style="font-size:10px"><a href="{2}">{1}</a> (<em>{3}</em>)</span>'
)

by_collection_tab <- cols_label(
  by_collection_tab,
  collection = html("<b>Collection</b>"),
  n_records  = html("<b>Records</b>"),
  n_species  = html("<b>Species</b>")
)

by_collection_tab <- data_color(
  by_collection_tab,
  vars(n_records),
  col_numeric(c("grey95", "grey50"), c(0L, by_collection$n_records))
)

data_color(
  by_collection_tab,
  vars(n_species),
  col_numeric(c("grey95", "grey50"), c(0L, by_collection$n_species))
)
```

```{r by-collection-plot, echo = FALSE, fig.align = "center", fig.width = 7, fig.height = 10}
informal_groups <- trimws(
  grep("^ ", finbif:::informal_group$tree, invert = TRUE, value = TRUE)
)

informal_groups <- tools::toTitleCase(informal_groups)

informal_groups_n <- setNames(
  seq_along(informal_groups), informal_groups
)

for (i in informal_groups) {
  filter <- c(list(informal_group = i), last_week, permissive)
  informal_groups_n[[i]] <- fb_occurrence(filter = filter, count_only = TRUE)
}

names(informal_groups_n) <- tools::toTitleCase(informal_groups)

obs <- fb_occurrence(
  filter = c(list(country = "Finland"), last_week, permissive),
  select = c("lon_wgs84", "lat_wgs84", "date_time", "source_id", "collection"),
  n = n, quiet = TRUE
)

obs$source <- source_names[obs$source_id]
obs$collection = tools::toTitleCase(obs$collection)

ggplot(
  filter(obs, !is.na(date_time)),
  aes(
    x = reorder(collection, desc(collection)), y = date_time,
    fill = source, color = source
  )
) +
geom_boxplot(outlier.size = 1L, varwidth = TRUE) +
scale_y_datetime(
  expand = expansion(c(.1, 0)), position = "right", date_breaks = "10 years",
  date_labels = "%Y"
) +
scale_x_discrete(labels = wrap_format(40L), position = "top") +
scale_fill_brewer(palette = "Dark2") +
scale_color_brewer(palette = "Dark2") +
coord_flip() +
facet_grid(
  rows = vars(source), scales = "free_y", switch = "y", space = "free_y",
  labeller = label_wrap_gen(20L)
) +
theme_minimal() +
theme(
  strip.text.y.left = element_text(angle = 0L, size = 10L),
  strip.placement = "outside",
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  axis.text.y = element_text(size = 7L),
  axis.text.x = element_text(angle = 45L, hjust = 0L, size = 11L),
  legend.position = "none",
  panel.grid.major.y = element_blank()
)
```

```{r by-group, echo = FALSE, fig.align = "center", fig.width = 7}
m <- rep(0L, 4L)

par(mfrow = c(1L, 2L), oma = m)

par(mar = c(5L, 11L, 1L, 1L))
barplot(rev(informal_groups_n), las = 2L, border = NA, horiz = TRUE)

par(mar = m)
image(
  hist_xy(
    obs[c("lon_wgs84", "lat_wgs84")], breaks_xy(finland_map$bbox, size = .25)
  ),
  asp    = 2.4,
  breaks = 2^seq(0L, 12L),
  col    = hcl.colors(12L, rev = TRUE),
  xlab   = NA,
  ylab   = NA,
  axes = FALSE,
  bty = "n"
)

legend(
  "bottomleft",
  inset  = c(0, .01),
  legend = expression(2^12, "", "", 2^6, "", "", 2^0),
  fill   = hcl.colors(7L),
  border = NA,
  bty    = "n",
  adj    = c(0, 0.25),
  x.intersp = .2,
  y.intersp = .5
)

polygon(finland_map$vertices, lwd = .2)
```
<br>

**** 
<br>

## Latest Images <br>
```{r images, include = FALSE}
n_imgs <- 0L
images <- list(url = NULL, cap = NULL, scap = NULL)

while (n_imgs < 4L) {
  records <- fb_occurrence(
    filter = c(list(has_record_media = TRUE), last_week, permissive),
    select = c(
      "record_media_type", "record_media_url", "scientific_name", "record_id",
      "scientific_name_italicised"
    ),
    n = 1, sample = TRUE, cache = FALSE
  )
  is_img <- records[["record_media_type"]][[1L]] == "IMAGE"
  urls <- records[["record_media_url"]][[1L]][is_img]
  if (identical(length(urls), 0L) || is.na(records[["scientific_name"]])) next
  n_imgs <- n_imgs + 1L
  images$url[[n_imgs]]  <- sample(urls, 1L)
  images$cap[[n_imgs]] <- paste0("<i>", records[["scientific_name"]], "</i>")
  i <- records[["scientific_name_italicised"]]
  i <- if (is.na(i) || !i) "span" else "i"
  images$cap[[n_imgs]] <- sprintf(
    '<%1$s>%2$s</%1$s> (<a href="%3$s">%3$s</a>)',
    i, records[["scientific_name"]], records[["record_id"]]
  )
}

img_obj <- lapply(images$url, image_read)
img_obj <- lapply(img_obj, image_resize, geometry_size_pixels(width = 500L))
```

```{r image-plotting, echo = FALSE, fig.align = "center", fig.cap = images$cap, fig.width = 3.5}
img_obj[[1L]]
img_obj[[2L]]
img_obj[[3L]]
img_obj[[4L]]
```

****
<br>

## Latest Rarities <br>
```{r rarities, echo = FALSE}
i <- 0L
n <- 0L

while(n < 10L && i < 50L) {
  i <- i + 1L
  n <- fb_occurrence(
    filter = c(
      last_week,
      permissive,
      list(n_total_records_finland_max = i, country = "Finland", finnish = TRUE)
    ),
    count_only = TRUE
  )
  Sys.sleep(1L)
}

rarities <- fb_occurrence(
  filter = c(
    last_week,
    permissive,
    list(n_total_records_finland_max = i, country = "Finland", finnish = TRUE)
  ),
  select = c(
    "date_time", "abundance", "scientific_name", "municipality", "record_id",
    "taxon_id", "scientific_name_italicised"
  )
)

rarities <- mutate(
  rarities,
  scientific_name_italicised = ifelse(scientific_name_italicised, "em", "span"),
  date_time = format(date_time, "%b %e, %Y")
)

rarities <- gt(rarities)

rarities <- tab_header(
  rarities,
  title = paste(
    "These taxa have less than", n + 1L, "observations recorded in FinBIF"
  )
)

rarities <- cols_merge(
  rarities,
  vars(date_time, record_id), pattern = '<a href="{2}">{1}</a>'
)

rarities <- cols_merge(
  rarities,
  vars(scientific_name, taxon_id, scientific_name_italicised),
  pattern = '<{3}><a href="{2}">{1}</a></{3}>'
)

cols_label(
  rarities,
  date_time = html("<b>Date</b>"),
  scientific_name  = html("<b>Taxa</b>"),
  abundance  = html("<b>Count</b>"),
  municipality = html("<b>Municipality</b>")
)
```

****
<br>

## Latest Publications <br>
```{r publications, echo = FALSE, results = "asis"}
format_authors <- function(x) {
  authors <- lapply(
    x$author,
    function(x) paste(x$given, trimws(paste(x$`dropping-particle`, x$family)))
  )
  authors <- lapply(authors, paste, collapse = " ")
  authors <- rev(authors)
  for (i in seq_along(authors)) {
    if (i == 2L) authors[[i]] <- paste0(authors[[i]], " & ")
    if (i > 2L) authors[[i]] <- paste0(authors[[i]], ", ")
  }
  authors <- rev(authors)
  do.call(paste0, authors)
}
bib <- jsonlite::read_json("https://luomus.github.io/finbif-bib/bib-data.json")
date_added <- vapply(bib, getElement, character(1L), "dateAdded")
bib <- head(sample(bib[as_date(date_added) >= as_date(last_week[[1L]])]))
for(i in bib)
cat(
  '<article itemscope="" itemtype="https://schema.org/ScholarlyArticle"',
  ' class="publication"><header><p itemprop="name"',
  ' class="publication-title"><a href="',
  i$url, '" target="_blank">', i$title,
  '</a></p><p class="authors">', format_authors(i),
  '</p><p><span itemprop="isPartOf" itemscope=""',
  ' itemtype="https://schema.org/Periodical"><span itemprop="name"',
  ' class="publication-journal"><i>',
  i$containerTitle,
  '</i></span></span><span>&nbsp;</span><span itemprop="isPartOf" itemscope=""',
  ' itemtype="https://schema.org/PublicationVolume"><span',
  ' itemprop="volumeNumber" class="publication-volume"><b>',
  i$volume,
  '</b></span></span><span>&nbsp;</span><span itemprop="pages"',
  ' class="publication-pages">',
  i$page, '</span><span>&nbsp;</span><time itemprop="datePublished">(',
  i$month,' ', i$year,
  ')</time></p></header></article>\n<hr>\n',
  sep = ""
)
```
